---
title: "HW 3 (Replication of Porter 1983)"
output: html_notebook
---

In this homework, we are asked to replicate the results of "A study of cartel stability: the Joint Executive Committee" paper by R. H. Porter (1983, The Bell Journal of Economics).

We use the <a href="https://ocw.mit.edu/courses/economics/14-271-industrial-organization-i-fall-2013/assignments/Porter.prn"> dataset </a> provided by Prof. Glenn Ellison.

I will use <b>R</b> for this homework.

Let's start.

Step 0: I'll import the necessary R packages that I'll use later.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(AER)
library(stats)
```
<br></br>

Step 1: I import the dataset into R, and name the columns.
```{r}
data <- read.table("https://ocw.mit.edu/courses/economics/14-271-industrial-organization-i-fall-2013/assignments/Porter.prn", header = TRUE)

```
<br></br>

Let's first see how the data looks.
```{r}
glimpse(data)
summary(data)
```

To start with, let's do a simple OLS regression of ``` log(quantity)``` on a constant, ```
log(price)```, ```lakes```, and all the seasonal dummies (except one, since we have already included a constant. I'll exclude the first one). 






































